Analysis2

### 2023-10-28 Lahman Analysis/Rough Analysis 2/Analysis2.qmd
Improve model with SB and model assessments
I added some more graphical evaluations of the model with SB. I have used proportion of home runs to evaluation the model.

ggplot(data = Batting_summary_SB %>% filter(SB < 50)) +
  geom_pointrange(mapping = aes(x = SB, y = logit(prop), ymin = logit(prop - se),
                                ymax = logit(prop + se))) +
  geom_smooth(method = "lm", mapping = aes(x = SB, y = logit(prop)))
```
To prepare to fit the model, we summarize the data, grouping player-seasons by year, player age, and number of stolen bases. (As in other examples, this grouping will not affect the likelihood, and it seems to speed up the work of Stan considerably.)

```{r}
sum_all_years_SB <- Batting %>% group_by(yearID, Age, SB) %>%
  summarize(HR = sum(HR), AB = sum(AB)) %>% ungroup
```

@@

@@

@@

(Here is an RStan demo with some inspirational code: [http://avehtari.github.io/BDA_R_demos/demos_rstan/ppc/poisson-ppc.html])

@@

```{r}
ppc_hist(sum_all_years_SB %>% pull(HR), HR_rep[1:8, ], binwidth = 1)
```

Yuck... A poor fit. The model doesn't reproduce low numbers of home runs as we see in the original data set.

```{r}
ppc_dens_overlay(sum_all_years_SB %>% pull(HR), HR_rep[1:50, ])
```
This family of density plots shows the same thing. A basic check of the model (histogram of home-run counts in our groups) shows a mismatch.

The situation reminds me of the pictures in Section 26.2 of the Stan User's Guide [https://mc-stan.org/docs/2_26/stan-users-guide/plotting-multiples.html]. In that section, the authors show an overdispersed count variable (negative binomial) fit with a binomial model. There is more on the same example in Section 26.3, using a test statistic and the p-value idea to quantify the poor fit.

>>>>

ggplot(data = Batting_summary_SB %>% filter(SB < 50)) +
  geom_pointrange(mapping = aes(x = SB, y = logit(prop), ymin = logit(prop - se),
                                ymax = logit(prop + se))) +
  geom_smooth(method = "lm", mapping = aes(x = SB, y = logit(prop)))
```
Since we will use both stolen bases and age in the model, we can make the same sort of graph after grouping players by age. Let's look first at the distribution of age.

```{r}
ggplot(data = Batting) +
  geom_histogram(mapping = aes(x = Age), binwidth = 1)
```
(What a lovely histogram!)

We will split players into three roughly equally sized age groups, determined by quantiles.

```{r}
Batting %>% summarize(quantile(Age, 0.33), quantile(Age, 0.67))
```
Our three groups are "26 or younger", "27 to 30", and "over 30".

```{r}
Batting <- Batting %>% mutate(age_group = case_when(
  Age <= 26 ~ "young",
  Age <= 30 ~ "middle",
  Age > 30 ~ "old"
))
```

We repeat our graphical exploration, faceting by age group. The general pattern (negative association between SB and home-run probability) persists across age groups. We note, though, that the pattern looks a bit different for the youngest players

```{r}
Batting_summary_age_SB <- Batting %>% group_by(age_group, SB) %>%
  summarize(HR = sum(HR), AB = sum(AB)) %>%
  mutate(prop = HR / AB, se = sqrt(prop*(1-prop)/AB))

ggplot(data = Batting_summary_age_SB) +
  geom_pointrange(mapping = aes(x = SB, y = prop, ymin = prop - se,
                                ymax = prop + se)) +
  facet_wrap(~age_group)
```

```{r}
ggplot(data = Batting_summary_age_SB) +
  geom_pointrange(mapping = aes(x = SB, y = logit(prop), ymin = prop - se,
                                ymax = prop + se)) +
  facet_wrap(~age_group)
```

To prepare to fit the model, we summarize the data, grouping player-seasons by year, player age, and number of stolen bases. (As in other examples, this grouping will not affect the likelihood, and it seems to speed up the work of Stan considerably.)

```{r}
sum_all_years_SB <- Batting %>% group_by(yearID, Age, SB) %>%
  summarize(HR = sum(HR), AB = sum(AB)) %>% mutate(prop = HR / AB) %>% ungroup
```

@@

Here is the actual distribution of HR proportions in the data. We shouldn't be trying to fit the data with the prior, but we can see how far off things are from this second histogram.

```{r}
ggplot(data = sum_all_years_SB) +
  geom_histogram(mapping = aes(x = prop))
```

@@

We can repeat setting stronger priors.

```{r}
prior24b2 <- stan(
  file = "HR24b_prior_sim2.stan",
  data = data24b,
  algorithm = "Fixed_param",
  iter = 1,
  chains = 1
)
```

```{r}
HR_sim2 <- (rstan :: extract)(prior24b2)$HR[1,]

sum_all_years_SB2 <- sum_all_years_SB %>% mutate(HR_sim = HR_sim2,
                                                prop_sim = HR_sim / AB)
```

```{r}
ggplot(data = sum_all_years_SB2) +
  geom_histogram(mapping = aes(x = prop_sim))
```

That prior distribution of simulated HR proportions looks better than what we got with the very weak priors on `beta1` and `c`. But perhaps these priors would be too strong...

@@

(Here is an RStan demo with some inspirational code for working with the replications: [http://avehtari.github.io/BDA_R_demos/demos_rstan/ppc/poisson-ppc.html]. Here is another vignette with examples of posterior predictive checks: [https://cran.r-project.org/web/packages/bayesplot/vignettes/graphical-ppcs.html])

A better way to do this would be to remove the rows in the data table for playes with 0 at-bats. We could then have Stan calculate proportions (without divide-by-0 errors). I got around those rows in an ugly way below.

@@

```{r}
ppc_hist(sum_all_years_SB %>% filter(AB > 0) %>% pull(prop) ,
         HR_rep[1:8, sum_all_years_SB$AB > 0] / (sum_all_years_SB %>% filter(AB > 0) %>% pull(AB))) +
  xlim(0, 0.3)
```

Yuck... A poor fit. The model replications are quite consistent---and don't look much like the original data. In the model, home-run percentages over 10% and as high as 20% sometimes occur. These never happen in the real data. The distribution of proportions in the simulated data has mode at close to 0. In the real data, the model appears closer to 3%.

```{r}
ppc_dens_overlay(sum_all_years_SB %>% filter(AB > 0) %>% pull(prop) ,
         HR_rep[1:50, sum_all_years_SB$AB > 0] / (sum_all_years_SB %>% filter(AB > 0) %>% pull(AB))) +
  xlim(0, 0.3)
```
This family of density plots shows the same thing as the histograms. The data from the model are quite consistent---and quite different from the original data.
